Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-01 _tex/paper2.tex
Built site for gh-pages

For instance, a motivating example that illustrates these concepts
throughout the document could involve researchers aiming to answer the
question: ``To what extent do different teaching methods influence
students' ability to produce high-quality written texts?'' To
investigate this, they could design a CJ experiment by randomly
assigning students to two groups, each receiving a different teaching
method. Judges would then compare pairs of students' written texts,
producing a dichotomous outcome that reflects the relative quality of
each text. Based on this setup, researchers can reformulate the research
question as the estimand: ``\emph{On average}, is there a difference in
the ability to produce high-quality written texts between the two groups
of students?''. Following current CJ practices, researchers would then
rely on estimates from the BTL model, or its transformations, to
approximate this estimand.

However, Section~\ref{sec-theory-issues} presents compelling evidence
that the BTL model has several statistical and measurement limitations.
These limitations hinder the model's ability to identify distinct
estimands relevant to CJ inquiries, including the one from the
motivating example. Identification is crucial because it is a necessary
condition for ensuring consistent estimators. Without identification,
researchers cannot achieve \emph{consistency} even with ``infinite'' and
error-free data. In other words, the estimates derived from the
estimator would fail to converge to the ``true'' value of the estimand
as the data size approaches infinity \citep{Everitt_et_al_2010}. As a
result, deriving meaningful insights from finite data also becomes
unattainable \citep[pp.~5]{Schuessler_et_al_2023}.

@@

As a result, this section addresses the identified limitations by
extending Thurstone's general form (see Table~\ref{tbl-thurstone_cases})
through a systematic and integrated approach based on the structural
approach to causal inference. Specifically, it leverages the
capabilities of this approach to formalize the combination of the core
theoretical principles outlined in Section~\ref{sec-thurstone_theory}
with key assessment design features relevant to CJ experiments, such as
the selection of judges, stimuli, and comparisons. The extended model
clarifies the interactions among all actors and processes involved in CJ
experiments. Furthermore, it allows researchers to shift the current
comparative data analysis paradigm from accepting the model's
assumptions as given to testing these assumptions for specific
comparative data. For details on these assumptions, refer to
Section~\ref{sec-theory-issues}. Accordingly,
Section~\ref{sec-theory-theoretical_P} incorporates the theoretical
principles into what the study calls the \emph{conceptual-population
model}. Meanwhile, Section~\ref{sec-theory-theoretical_SC} integrates
the assessment design features into what the study terms the
\emph{sample-comparison model}.

@@

Before integrating the theoretical principles from
Section~\ref{sec-thurstone_theory}, the study defines SCMs and DAGs.
SCMs are formal mathematical models characterized by a set of
\emph{endogenous} variables \(V\), a set of \emph{exogenous} variables
\(E\), and a set of functions \(F\)
\citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous variables are those
whose causal mechanisms a researcher chooses to model \citep{Neal_2020}.
In contrast, exogenous variables represent \emph{errors} or
\emph{disturbances} arising from omitted factors that the investigator
chooses not to model explicitly \citep[pp.~27,68]{Pearl_2009}. Lastly,
the functions, referred to as \emph{structural equations}, express the
endogenous variables as non-parametric functions of other variables.
These functions use the symbol `\(:=\)' to denote the asymmetrical
causal dependence between variables and the symbol `\(\:\bot\:\)' to
represent \emph{d-separation}, a concept akin to (conditional)
independence.

Figure~\ref{fig-cj04} illustrates the preliminary integration of the
theoretical principles underlying the motivating example. Specifically,
SCM~\ref{fig-cj04_scm} displays the relationship between the outcome of
the CJ experiment, \(O_{abhijk}\), to several other variables. Here, the
indices \(\{a,b\}\) represent the compared texts, \(\{h,i\}\) label the
students who wrote these texts, \(j\) denotes the judge, and \(k\)
accounts for cases where the judges perform multiple comparisons of the
same texts written by the same students, i.e., the comparison number for
a \emph{repeated measures design} \citep[pp.~366-376]{Lawson_2015}. The
other variables related to the outcome include the discriminal
difference \((D_{abhijk})\), the perceived discriminal processes
\((PT_{ahjk}, PT_{bijk})\), the ``true'' discriminal processes
\((T_{ahjk}, T_{bijk})\), and the judges' biases \((B_{kj})\). Together
with the outcome, these variables form the preliminary set of endogenous
variables
\(V = \{O_{abhijk}, D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}\).
Finally, the preliminary set of structural equations
\(F = \{f_{O},f_{D},f_{PT}\}\) defines the dependencies among these
variables.

@@

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
consisting of nodes connected by edges, where the nodes represent random
variables. The term \emph{directed} means that the edges extend from one
node to another, with arrows indicating the direction of causal
influence. The term \emph{acyclic} implies that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. While the \emph{standard representation} of
DAGs typically omits exogenous variables for simplicity, the
\emph{magnified representation} includes them, offering the advantage of
highlighting potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}.

DAG~\ref{fig-cj04_dag} displays the preliminary DAG associated with
SCM~\ref{fig-cj04_scm}. The graph shows the direction of causal
influences among variables, reflecting the assumptions of Thurstone's
theory. It also distinguishes between observable endogenous variables,
such as the outcome \(O_{abhijk}\), and variables considered latent
according to the theory and practice of CJ, such as the discriminal
difference \((D_{abhijk})\), the perceived discriminal processes
\((PT_{ahjk}, PT_{bijk})\), the ``true'' discriminal processes
\((T_{ahjk}, T_{bijk})\), and the judges' biases \((B_{kj})\).

@@

Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
display the five fundamental building blocks for SCMs and DAGs. The left
panels of the figures show the formal mathematical models, represented
by the SCMs, defined in terms of a set of \emph{endogenous} variables
\(X=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
\(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions
\(F=\{f_{X1},f_{X2},f_{X3}\}\) \citep{Pearl_2009, Cinelli_et_al_2020}.
Endogenous variables are those whose causal mechanisms a researcher
chooses to model \citep{Neal_2020}. In contrast, exogenous variables
represent \emph{errors} or \emph{disturbances} arising from omitted
factors that the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence of the
variables and the symbol `\(\:\bot\:\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

>>>>

A motivating example helps to clarify these concepts. The example will
appear throughout the document to illustrate various aspects of the
proposed extension. In this scenario, researchers aim to determine: ``To
what extent do different teaching methods influence students' ability to
produce high-quality written texts?'' To investigate this, they could
design a CJ experiment by randomly assigning students (individuals) to
two groups, each receiving a different teaching method. Judges would
then compare pairs of students' written texts (stimuli), producing a
dichotomous outcome that reflects the relative quality of each text
(trait). Based on this setup, researchers could reformulate the research
question as the estimand: ``\emph{On average}, is there a difference in
the ability to produce high-quality written texts between the two groups
of students?''. Following current CJ practices, researchers would then
rely on estimates from the BTL model, or its transformations, to
approximate this estimand.

However, Section~\ref{sec-theory-issues} presents compelling evidence
that the BTL model has several statistical and measurement limitations.
These limitations hinder the model's ability to identify various
estimands relevant to CJ inquiries, including the one from the
motivating example. Identification is crucial because it is a necessary
condition for ensuring consistent estimators. Without identification,
\emph{consistency} --the property of an estimator whose estimates
converge to the ``true'' value of the estimand as the data size
approaches infinity \citep{Everitt_et_al_2010}-- cannot be achieved,
even with ``infinite'' and error-free data. As a result, deriving
meaningful insights from finite data also becomes unattainable
\citep[pp.~5]{Schuessler_et_al_2023}.

@@

As a result, this section addresses the identified limitations by
extending Thurstone's general form (see Table~\ref{tbl-thurstone_cases})
through a systematic and integrated approach based on the structural
approach to causal inference. Specifically, it leverages the
capabilities of this approach to formalize the combination of the core
theoretical principles outlined in Section~\ref{sec-thurstone_theory}
with key assessment design features relevant to CJ experiments, such as
the selection of judges, stimuli, and comparisons.

In addition to addressing the identified limitations, the extended model
offers two key advantages. First, it clarifies the interactions among
all actors and processes involved in CJ experiments. Second, it shifts
the current data analysis paradigm from passively accepting CJ's model
assumptions to actively testing them with the specific data under
analysis. For more details on these assumptions, refer to
Section~\ref{sec-theory-issues}.

Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
theoretical principles into what the study calls the
\emph{conceptual-population model}. This model assumes that researchers
have access to a conceptual population of comparison data, meaning data
representing all repeated comparisons made by every available judge for
each pair of stimuli produced by each pair of individuals in the
population, thus its name. Conversely,
Section~\ref{sec-theory-theoretical_SC} integrates the assessment design
features into what the study calls the \emph{sample-comparison model}.
This model assumes a more realistic scenario where researchers only have
access to a sample of judges, individuals, stimuli, and comparisons from
the conceptual population.

@@

Before incorporating the theoretical principles into the
\emph{conceptual-population model}, the study defines SCMs, which are
formal mathematical models characterized by a set of \emph{endogenous}
variables \(V\), a set of \emph{exogenous} variables \(E\), and a set of
functions \(F\) \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other endogenous and exogenous variables.
These functions use the symbol `\(:=\)' to denote the asymmetrical
causal dependence between variables and the symbol `\(\:\bot\:\)' to
represent \emph{d-separation}, a concept akin to (conditional)
independence.

Figure~\ref{fig-cj04} illustrates the preliminary integration of
theoretical principles into the conceptual-population model.
Specifically, Figure~\ref{fig-cj04_scm} shows the SCM that underlies the
motivating example. This SCM maps the relationship between the outcome
of the CJ experiment \((O_{abhijk})\) to several other variables. In
this model, the indices \(\{a,b\}\) represent the compared texts
(stimuli), with \(a \neq b\). The indices \(\{h,i\}\) label the students
who wrote the texts (individuals), where \(h = i\) is possible. The
index \(j\) denotes the judge, and \(k\) denotes the comparison number,
which accounts for experimental cases where judges perform multiple
comparisons of the same texts written by the same students, i.e., a
\emph{repeated measures design} \citep[pp.~366-376]{Lawson_2015}.
Finally, the SCM presents the preliminary set of structural equations
\(F = \{f_{O},f_{D},f_{PT}\}\), which define the non-parametric
dependencies among these variables.

Note that the SCM~\ref{fig-cj04_scm} incorporates not only the variables
described by Thurstone's theory, such as the discriminal difference
\((D_{abhijk})\) and the texts' \emph{true} discriminal processes
\((T_{ahjk}, T_{bijk})\) (refer to Section~\ref{sec-thurstone_theory}),
but also innovates by incorporating additional variables central to CJ
experiments, such as the texts' \emph{perceived} discriminal processes
\((PT_{ahjk}, PT_{bijk})\) and the judges' perceptions \((B_{kj})\).
Together with the outcome, these variables constitute the preliminary
set of endogenous variables
\(V = \{O_{abhijk}, D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}\).

By incorporating the texts' perceived discriminal processes into the
conceptual-population model, this study considers a crucial aspect of CJ
experiments that previous literature has overlooked. It recognizes that
the texts' true discriminal processes become an observable outcome only
through the judges' perceptions. In other words, researchers observe the
CJ outcome only when both variables create a \emph{pure interaction
effect}, meaning that neither variable has a direct effect on the
outcome, but their interaction produces an effect
\citep{Attia_et_al_2022}.

Conversely, based on the evidence presented in
Section~\ref{sec-theory-issue1b}, the study assumes that other relevant
stimulus characteristics influence the judges' perceptions, leading to
biases that resist cancellation and ultimately introduce dependencies
between the stimuli. To address this, the study incorporates the judges'
perceptions into the conceptual-population model, following the
recommendations of \citet{Andrich_1978} and \citet{Wainer_et_al_1978}.

@@

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
consisting of nodes connected by edges, where the nodes represent random
variables. The term \emph{directed} means that the edges extend from one
node to another, with arrows indicating the direction of causal
influence. The term \emph{acyclic} implies that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. While the \emph{standard representation} of
DAGs typically omits exogenous variables for simplicity, the
\emph{magnified representation} includes them, offering the advantage of
highlighting potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}. Thus, this study uses the magnified
representation.

Figure~\ref{fig-cj04_dag} displays the DAG associated with
SCM~\ref{fig-cj04_scm}. The graph shows the direction of causal
influences among variables, reflecting the assumptions of Thurstone's
theory and CJ experiments. It also distinguishes between observable
endogenous variables, such as the outcome \(O_{abhijk}\), and variables
considered latent according to the theory and practice of CJ, such as
the discriminal difference, the perceived discriminal processes, the
true discriminal processes, and the judges' perceptions.

It is easy to intuit that describing the CJ system by assigning an index
and a variable to each discriminal process could quickly become
intractable and cumbersome due to the large size of the DAG and the
repetition of its components. Therefore, to maintain generality, the
study groups pairs of discriminal processes with different indices into
matrices that contain these pairs. First, the study defines the vectors
\(U_{A} \in \mathbb{N}^{A}\), \(U_{I} \in \mathbb{N}^{I}\) containing
the indices of the stimuli and individuals, where:

\[
U_{A} = [1, \dots, a, \dots, b, \dots, A] ;\quad U_{I} = [1, \dots, a, \dots, b, \dots, I]
\]

Then, the study defines the matrices of stimuli and individual
comparisons, \(P \in \mathbb{N}^{m \times 2}\) and
\(R \in \mathbb{N}^{n \times 2}\), respectively. Here, each row of \(P\)
represents a pairing \(\{a,b\}\) of stimuli indices and each row of
\(R\) represents a pairing \(\{h,i\}\) of individual indices, i.e.,

\[
P = \{(a, b) \mid a,b \in U_{A}, a \neq b\} ;\quad R = \{(h, i) \mid h,i \in U_{I}\}
\]

Here, \(m\) and \(n\) represents the number of unique pairwise
comparisons between stimuli, and unique pairwise comparison per
individuals, which corresponds to the number of ways to choose two
elements from \(U_{A}\) and \(U_{I}\), typically given by:

\[
m = {A \choose 2} = \frac{A!}{2!(A-2)!} ;\quad n = {I \choose 2} = \frac{I!}{2!(I-2)!}
\]

@@

Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
display the five fundamental building blocks for SCMs and DAGs. The left
panels of the figures show the formal mathematical models, represented
by the SCMs, defined in terms of a set of \emph{endogenous} variables
\(V=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
\(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions
\(F=\{f_{X1},f_{X2},f_{X3}\}\) \citep{Pearl_2009, Cinelli_et_al_2020}.
Endogenous variables are those whose causal mechanisms a researcher
chooses to model \citep{Neal_2020}. In contrast, exogenous variables
represent \emph{errors} or \emph{disturbances} arising from omitted
factors that the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence of the
variables and the symbol `\(\:\bot\:\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.
